---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2021"
subtitle: "Assignment 2 - Due date 02/05/21"
author: "Nick Valby"
output: pdf_document
geometry: margin=1.5cm
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is change "Student Name" on line 4 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp21.Rmd"). Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r include=FALSE}
#Load/install required package here

#install.packages("forecast")
#install.packages("tseries")
#install.packages("dplyr")
#install.packages("lubridate")
#install.packages("readxl")


#Load packages
library("forecast")
library("tseries")
library("dplyr")
library("lubridate")
library("readxl")

```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the January 2021 Monthly Energy Review. The spreadsheet is ready to be used. Use the command $read.table()$ to import the data in R or $panda.read\_excel()$ in Python (note that you will need to import pandas package). }

```{r include=TRUE}
#Importing data set

energy_data <- read_excel("../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", col_names = FALSE, sheet = "Monthly Data", skip = 12) 

#Organizing the data 
energy_data_names <- read_excel("../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", col_names = FALSE, sheet = "Monthly Data", skip = 10, n_max = 1)

#Naming the columns
colnames(energy_data) <- energy_data_names

head(energy_data)

```

<br />
<br />

## Question 1
**You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.**

\medskip
\medskip

Here is the data frame structure with the month/year and three columns:
```{r include=TRUE}

energy_data <- select(energy_data, "Month", "Total Biomass Energy Production", "Total Renewable Energy Production", "Hydroelectric Power Consumption")

head(energy_data)

```
<br />
<br />

## Question 2
**Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().**

\medskip
\medskip

**Here is the data frame in a time series format starting in January 1973:**

\medskip

```{r include=TRUE}
#Transforming the data into a time series:
ts_energy_data <- ts(energy_data[,2:4],frequency = 12, start = 1973)

#Here is the data in a time series:
head(ts_energy_data)
```

<br />
<br />

\medskip

## Question 3
**Compute mean and standard deviation for these three series.**

\medskip

**Means of the three time series:**

Total Biomass Energy Production: `r mean(ts_energy_data[,1])`

<br />
<br />

Total Renewable Energy Production: `r mean(ts_energy_data[,2])`

<br />
<br />

Hydroelectric Power Consumption: `r mean(ts_energy_data[,3])`

\medskip
\medskip

**Standard deviations of the three time series:**

Total Biomass Energy Production: `r sd(ts_energy_data[,1])`

<br />
<br />

Total Renewable Energy Production: `r sd(ts_energy_data[,2])`

<br />
<br />

Hydroelectric Power Consumption: `r sd(ts_energy_data[,3])`

<br />
<br />

```{r include=TRUE}
mean(ts_energy_data[,1])
mean(ts_energy_data[,2])
mean(ts_energy_data[,3])

sd(ts_energy_data[,1])
sd(ts_energy_data[,2])
sd(ts_energy_data[,3])

```

\newpage


## Question 4
**Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.**

\medskip

```{r include=TRUE}
plot(ts_energy_data[,1],
     ylab = "Trillions of Btu",
     main = "Total Biomass Energy Production",
     type = "l")
abline(h=270.69, col = "red")

```

\medskip


```{r include=TRUE}
plot(ts_energy_data[,2],
     ylab = "Trillions of Btu",
     main = "Total Renewable Energy Production",
     type = "l")
abline(h=572.73, col = "red")
```

<br />
<br />

\


```{r include=TRUE}

plot(ts_energy_data[,3],
     ylab = "Trillions of Btu",
     main = "Hydroelectric Power Consumption",
     type = "l")
abline(h=236.95, col = "red")

```

<br />
<br />


## Question 5
**Compute the correlation between these three series. Are they significantly correlated? Explain your answer.**

\medskip
\medskip

**The correlation between Total Biomass Energy Production and Total Renewable Energy Production is `r cor(ts_energy_data[,1],ts_energy_data[,2])` which means the two variables are positively correlated**

\medskip

**The correlation between Total Renewable Energy Production and Hydroelectric Power Consumption is `r cor(ts_energy_data[,2],ts_energy_data[,3])` which means the variables are negatively correlated**

\medskip

**The correlation between Hydroelectric Power Consumption and Total Biomass Energy Production is `r cor(ts_energy_data[,3],ts_energy_data[,1])` which means the variables are weakly correlated in a negative direction**

<br />
<br />
```{r include=TRUE}

cor(ts_energy_data[,1],ts_energy_data[,2])

cor(ts_energy_data[,2],ts_energy_data[,3])

cor(ts_energy_data[,3],ts_energy_data[,1])


```

\newpage


## Question 6
**Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?**

\medskip
\medskip

The plots below do not have the same behavior. The first two have similar downward trends over time, while the third is obviously seasonal/cyclical.

\medskip
\medskip

```{r include=TRUE}

acf(energy_data[,2], lag.max = 40, main = "ACF of Total Biomass Energy Production")
```

\medskip

```{r include=TRUE}
acf(energy_data[,3], lag.max = 40, main = "ACF of Total Renewable Energy Production")
```

\medskip

```{r include=TRUE}
acf(energy_data[,4], lag.max = 40, main = "ACF of Hydroelectic Power Consumption")
```


\medskip
\medskip
\medskip

## Question 7
**Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?**


\medskip

The plots below are different from the ones in Q6 because these plots represent the difference between between a variable at two different times WITH the "white noise" removed.

\medskip
\medskip

```{r include=TRUE}
pacf(energy_data[,2], lag.max = 40, main = "PACF of Total Biomass Energy Production")
```

\medskip

```{r include=TRUE}
pacf(energy_data[,3], lag.max = 40, main = "PACF of Total Renewable Energy Production")
```

\medskip

```{r include=TRUE}
pacf(energy_data[,4], lag.max = 40, main = "PACF of Hydroelectic Power Consumption")
```





